Revealing transient strain in geodetic data with Gaussian process regression

### 2017-10-24 RevisedSubmission/hines_hetland_2017_gji.tex
did more

We seek a spatially and temporally dependent estimate of transient
strain rates. We consider transient strain rates to be any deviation
from the secular strain rates. We denote transient strain rates as
$\strain(p)$, where $p$ represents the ordered pair $(\pos,t)$, $\pos
= (x_\e,x_\n)$ are spatial coordinates, and $t$ is time. For
simplicity, we assume that the study region is sufficiently small that
$\pos$ can be considered a point in a 2-D Cartesian map projection
that is aligned with the cardinal directions. We determine $\strain$
by spatially and temporally differentiating estimates of transient
displacements, which are constrained by GNSS data.

@@

\begin{figure*}
\includegraphics{figures/prior_demo/prior-demo.pdf}
\caption{ 
Panel A) The covariance between $u_i(p)$ and $u_i(p')$,
$C_{u_i}(p,p')$, as a function of the spatial and temporal distance
between $p$ and $p'$. The covariance function $C_{u_i}$ is composed of
squared exponentials to describe the spatial and temporal covariance
of $u_i$ (eq. \ref{eq:ExampleCov}). Panel B) A single realization of
$u_i$ using the covariance function from Panel A. The realization is
really a function of three variables, $t$, $x_\mathrm{e}$, and
$x_\mathrm{n}$, although we only show its dependence on $t$ and one of
the spatial dimensions.
}    
\label{fig:PriorDemo}
\end{figure*}

GNSS data records transient displacements as well as other physical
and non-physical processes that we are not interested in. We first
consider component $i$ of a single GNSS displacement observation made
at station $j$, which is located at $\pos^{(j)}$, and time $t^{(k)}$.
We describe this observation, $d_i^{*(jk)}$, as a realization of the
random variable
\begin{align}\label{eq:Data}
\begin{split}
d_i^{(jk)} = &u_i\left(\pos^{(j)},t^{(k)}\right) + 
              \eta_i^{(jk)} + 
              a_i^{(j)} + b_i^{(j)}t^{(k)} + \\
             &c_i^{(j)}\sin\left(2 \pi t^{(k)}\right) +  
              e_i^{(j)}\cos\left(2 \pi t^{(k)}\right) + \\ 
             &f_i^{(j)}\sin\left(4 \pi t^{(k)}\right)  + 
              g_i^{(j)}\cos\left(4 \pi t^{(k)}\right), 
\end{split}
\end{align}
where $\eta_i^{(jk)}$ describes noise, $a_i^{(j)}$ is an offset that
is unique for each station, $b_i^{(j)}$ is the secular velocity at
$\pos^{(j)}$, and the sinusoids describe seasonal deformation (using
units of years for $t^{(k)}$). We then consider the column vector of
$n$ GNSS displacement observations, $\data^*_i$, where the subscript
indicates that we are still only considering component $i$ of
displacements. The observations are made at $m$ stations, and the
times and positions for each observation are described by the set
$\points$. $\data^*_i$ can be considered a realization of the random
vector $\data_i$, which is formed by evaluating eq. (\ref{eq:Data}) at
each point in $\points$. To write out $\data_i$ explicitly, we let
$\G$ be an $n \times 6m$ matrix consisting of the basis functions from
eq. (\ref{eq:Data}) (i.e., the linear trends and sinusoids for each
station) evaluated at each point in $\points$. The coefficients
corresponding to each basis function are collected into the column
vector $\mitbf{m}_i$. The noise for all the observations are described
by the column vector $\mitbf{\eta}_i$. We can then write $\data_i$ as
\begin{equation}\label{eq:DataVec}
\data_i = u_i(\points) + \mitbf{\eta}_i + \G\mitbf{m}_i,
\end{equation}
where the notation $u_i(\points)$ represents the column vector
$[u_i(p)]_{p \in \points}$.

@@

@@

@@

@@

@@

@@

We use our method to estimate transient strain rates in the Pacific
Northwest, and we are specifically interested in identifying strain
resulting from SSEs \citep[e.g.,][]{Dragert2001}. In the Pacific
Northwest, SSEs can be detected by monitoring for associated seismic
tremor \citep{Rogers2003}, which is actively being done by the Pacific
Northwest Seismic Network \citep{Wech2010}. We can thus compare the
tremor records to the transient strain rates estimated with GPR to
verify that we are indeed identifying strain from SSEs.

We use the daily displacement solutions generated by the Geodesy
Advancing Geosciences and EarthScope (GAGE) facility for continuous
GNSS stations \citep{Herring2016}. This data is publicly available and
can be found at www.unavco.org. We limit the dataset to the stations
and time ranges which are pertinent to the seven most recent SSEs in
the Puget Sound region. The earliest SSE considered in this study
began in August 2010, and the most recent SSE began in February 2017.
We use these most recent SSEs because the station coverage is
sufficiently dense for us to use maximum likelihood methods to
constrain prior models.  The positions of GNSS stations used to
estimate transient strain rates are shown in Figure \ref{fig:Context}.

@@

Before we determine the transient strain rates, we must establish a
prior for the transient displacements, $u$, and the noise, $\eta$. In
this section we discuss our choice for the noise covariance function
$C_\eta$. There have been numerous studies on temporally correlated
noise in GNSS data \citep[e.g.,][]{Zhang1997, Mao1999, Williams2004,
Langbein2008}. In these studies, temporally correlated noise was
described with some combination of Brownian motion, a first-order
Gauss-Markov (FOGM) process, and/or flicker noise. There is some
physical justification for using Brownian motion as a noise model
because it accurately describes the power spectrum of motion resulting
from instability in geodetic monuments \citep[e.g.,][]{Wyatt1982,
Wyatt1989}. Here we describe the time dependence of $\eta$ as a FOGM
process and consider $\eta$ to be spatially uncorrelated. A FOGM
process is a solution to the stochastic differential equation
\begin{equation}\label{eq:FOGMdef}
\dot{\eta}(t) + \alpha \eta(t) = \beta w(t),
\end{equation}
where $w(t)$ is white noise with unit variance. The FOGM process
degenerates to the commonly used Brownian motion noise model under the
condition that $\alpha=0$ and $\eta(0) = 0$. Our spatially
uncorrelated noise model that satisfies eq. (\ref{eq:FOGMdef}) is a
Gaussian process with zero mean and the covariance function
\begin{equation}\label{eq:FOGM}
C_\eta\left((\pos,t),(\pos',t')\right) = 
\begin{cases}
\frac{\beta^2}{2\alpha}\exp\left(-\alpha|t - t'|\right), &\pos = \pos' \\
0, &\pos \neq \pos'
\end{cases}. 
\end{equation}

>>>>

We seek a spatially and temporally dependent estimate of transient
strain rates. We consider transient strain rates to be any deviation
from the secular strain rates. We denote transient strain rates as
$\strain(p)$, where $p$ represents the ordered pair $(\pos,t)$, $\pos
= (x_\e,x_\n)$ are spatial coordinates, and $t$ is time. For
simplicity, we assume that the study region is sufficiently small that
$\pos$ can be considered a point in a 2-D Cartesian map projection
that is aligned with the cardinal directions. We determine $\strain$
by spatially and temporally differentiating estimates of transient
displacements, which we constrain with GNSS data.

@@

\begin{figure*}
\includegraphics{figures/prior_demo/prior-demo.pdf}
\caption{ 
Panel A) The covariance between $u_i(p)$ and $u_i(p')$,
$C_{u_i}(p,p')$, as a function of the spatial and temporal distance
between $p$ and $p'$. The covariance function $C_{u_i}$ is composed of
squared exponentials to describe the spatial and temporal covariance
of $u_i$ (eq. \ref{eq:ExampleCov}). Panel B) A single realization of
$u_i$ corresponding to the covariance function from Panel A. The
realization is a function of three variables, $t$, $x_\mathrm{e}$, and
$x_\mathrm{n}$, although we only show its dependence on $t$ and one of
the spatial dimensions.
}    
\label{fig:PriorDemo}
\end{figure*}

% define the components of a datum

GNSS data records transient displacements as well as other physical
and non-physical processes that we are not interested in. We first
consider component $i$ of a single GNSS displacement observation made
at station $j$, which is located at $\pos^{(j)}$, and time $t^{(k)}$.
We describe this observation, $d_i^{*(jk)}$, as a realization of the
random variable
\begin{align}\label{eq:Data}
\begin{split}
d_i^{(jk)} = &u_i\left(\pos^{(j)},t^{(k)}\right) + 
              \eta_i^{(jk)} + 
              a_i^{(j)} + b_i^{(j)}t^{(k)} + \\
             &c_i^{(j)}\sin\left(2 \pi t^{(k)}\right) +  
              e_i^{(j)}\cos\left(2 \pi t^{(k)}\right) + \\ 
             &f_i^{(j)}\sin\left(4 \pi t^{(k)}\right)  + 
              g_i^{(j)}\cos\left(4 \pi t^{(k)}\right), 
\end{split}
\end{align}
where $\eta_i^{(jk)}$ describes noise, $a_i^{(j)}$ is an offset that
is unique for each station, $b_i^{(j)}$ is the secular velocity at
$\pos^{(j)}$, and the sinusoids describe seasonal deformation (using
units of years for $t^{(k)}$).

% define the data vector

We then consider the column vector of $n$ GNSS displacement
observations, $\data^*_i$, where the subscript indicates that we are
still only considering component $i$ of displacements. The
observations are made at $m$ stations, and the times and positions for
each observation are described by the set $\points$. $\data^*_i$ can
be considered a realization of the random vector $\data_i$, which is
formed by evaluating eq. (\ref{eq:Data}) at each point in $\points$.
To write out $\data_i$ explicitly, we let $\G$ be an $n \times 6m$
matrix consisting of the basis functions from eq. (\ref{eq:Data})
(i.e., the linear trends and sinusoids for each station) evaluated at
each point in $\points$. The coefficients corresponding to each basis
function are collected into the column vector $\mitbf{m}_i$. The noise
for all the observations are described by the column vector
$\mitbf{\eta}_i$. We can then write $\data_i$ as
\begin{equation}\label{eq:DataVec}
\data_i = u_i(\points) + \mitbf{\eta}_i + \G\mitbf{m}_i,
\end{equation}
where the notation $u_i(\points)$ represents the column vector
$[u_i(p)]_{p \in \points}$.

% establishing the distribution of the parameters in d and d itself

@@

% Combining the data and prior to form posterior transient
% displacements

@@

% Noting that we ignore covariance between components

@@

% converting posterior transient displacements to posterior transient
% strain

@@

% Non-technical description of outlier detection methods and our
% outlier detection method

@@

% Acknowledging that jumps are not detected

@@

% Overview of what we are doing in this section

In this section we estimate transient strain rates in the Pacific
Northwest, and we are specifically interested in identifying transient
strain resulting from SSEs \citep[e.g.,][]{Dragert2001}. Before
estimating transient strain rates, we establish a noise model for GNSS
stations in this region, and we establish a prior Gaussian process to
describe displacements from SSEs. SSEs in the Pacific Northwest can be
detected by monitoring for associated seismic tremor
\citep{Rogers2003}, which is actively being done by the Pacific
Northwest Seismic Network \citep{Wech2010}. We can then compare the
tremor records to our estimated transient strain rates to verify that
we are indeed identifying strain from SSEs.

% Dataset used here

We use the daily displacement solutions for continuous GNSS stations
generated by the Geodesy Advancing Geosciences and EarthScope (GAGE)
Facility \citep{Herring2016}. We limit the dataset to the stations and
times that are pertinent to the seven most recent SSEs in the Puget
Sound region. The earliest SSE considered in this study began in
August 2010, and the most recent SSE began in February 2017. The
positions of GNSS stations used to estimate transient strain rates are
shown in Figure \ref{fig:Context}.

@@

% landing on the FOGM noise model

In this section we establish a noise model for the GNSS data. The
noise in GNSS data, specifically the temporally correlated noise, has
been thoroughly studied over the past two decades
\citep[e.g.,][]{Zhang1997, Mao1999, Williams2004, Langbein2008}. In
these studies, temporally correlated noise is described with some
combination of Brownian motion, a first-order Gauss-Markov (FOGM)
process, and/or flicker noise. There is some physical justification
for using Brownian motion as a noise model because it accurately
describes the power spectrum of motion resulting from instability in
geodetic monuments \citep[e.g.,][]{Wyatt1982, Wyatt1989}. In
particular, the power spectrums for Brownian motion and monument
motion are both proportional to $f^{-2}$, where $f$ is frequency.
Brownian motion (which is also known as a Weiner process), by
definition, contains a reference time at which the process begins.
Since there is no notion of when noise ``begins" in GNSS data, we do
not find Brownian motion to be an appropriate noise model. On the
other hand, a FOGM process also has a power spectrum that is
proportional to $f^{-2}$, above some corner frequency, and it has no
reference time (i.e., a FOGM process is stationary). In this study, we
describe the temporal covariance of GNSS noise with a FOGM process. We
do not explore spatial covariance in GNSS noise and simply assume that
it is spatially uncorrelated.

Here we
describe the time dependence of $\eta$ as a FOGM process and consider
$\eta$ to be spatially uncorrelated. A FOGM process is a solution to
the stochastic differential equation

\begin{equation}\label{eq:FOGMdef}
\dot{\eta}(t) + \alpha \eta(t) = \beta w(t),
\end{equation}
where $w(t)$ is white noise with unit variance. The FOGM process
degenerates to the commonly used Brownian motion noise model under the
condition that $\alpha=0$ and $\eta(0) = 0$. Our spatially
uncorrelated noise model that satisfies eq. (\ref{eq:FOGMdef}) is a
Gaussian process with zero mean and the covariance function
\begin{equation}\label{eq:FOGM}
C_\eta\left((\pos,t),(\pos',t')\right) = 
\begin{cases}
\frac{\beta^2}{2\alpha}\exp\left(-\alpha|t - t'|\right), &\pos = \pos' \\
0, &\pos \neq \pos'
\end{cases}. 
\end{equation}
